Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
adding section of 7

\note{for reference notes are in \url{http://tinyurl.com/l4tc6a4}}
%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit

@@

\subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
\note{writing in progress by Frank Seinstra - expected by 22 March 2015}

sub-breakout: \url{http://tinyurl.com/mx58tmx}
%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit

>>>>

\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit

@@

\subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
\note{by Frank Seinstra}

\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit

At the start of the discussion, the breakout group brought forth the
important observation of the wide disparity in commonly accepted habits of
publication in different research fields.
In domains which have, historically, relied on large groups of researchers
collaborating towards a common goal (e.g., high-energy physics,
astronomy), publications often have tens or even hundreds of co-authors (with
some papers in experimental particle physics having over 3000.)
In other domains, the number of co-authors is typically much smaller,
with, in some cases, even a preference for single-author papers.
Similarly, the various platforms for publication are valued differently in
different domains.
Most commonly, publications in peer-reviewed scientific journals are regarded
as the most important and most impactful.
However, in certain domains, especially in Computer Science, many researchers
typically regard conference proceedings as their prime publication target.
It is often suggested that this difference is due to the rapid developments
in ICT science, a pace that can not be upheld by traditional peer-reviewed
journals.
Whatever the causes, any useful taxonomy of contributorship or guideline for
tenure review should take such differences into account.

Despite these differences, and despite the fact that software often
has taken the role of a proper, albeit less tangible, scientific research
instrument, neither the software nor its creators are commonly credited as part
of a scientific publication.
The group acknowledged the need for more recognition for the creators of such
software instruments, and indicated a number of possible pathways.
First and foremost, domain scientists must be made aware of the important role
of software, and include the developers as co-authors of papers.
A second approach is to fully embrace an open badging infrastructure (such as
Mozilla's Open Badges), where a {\em badge\/} is a free, transferrable, 
evidence-based indicator of an accomplishment, skill, quality, or interest.
A third approach is for the scientific community to support the increasing
momentum of peer-reviewed journals specialized in the open source/open
access publication of scientific research software, such as Computer Physics
Communication, F1000 Research, Journal of Open Research Software, and SoftwareX.

Recognizing publication of research software as a proper scientific
contribution raises several important but currently unsolved questions,
however.
For example, is the number of users of the software a relevant measure of
impact?
What standards of coding quality must be followed in order to justify
publication and hence recognition?
Should the release of a new version of the software be eligible for a new
publication; if so, under what conditions?
And above all: should software publications be valued in the same way as
traditional scientific publications?
Or is there a need for new measures of productivity and impact?

In part, the answers will come from the scientific community at large, as a
natural consequence of growing awareness and mindset change.
Some of the answers, however, also should be based on decades of experience
in (and developing standards for) implementing, maintaining, refactoring,
documenting, testing, and deploying software instruments in scientific
research.
Care should be taken, however, not to impose such standards for all domains
in equal ways right from the start.
Forerunners should serve as an example, but should not scare away domains that
have based their progress on much less advanced methods of {\em software
carpentry}.
Nevertheless, proper guidelines are needed, which eventually should be followed
across all domains.
The group also recognized that funding bodies, universities, and publishers
eventually should demand that research projects follow such guidelines, and to
implement a proper software sustainability plan.

To enable a form of standardized crediting for developers of research
software, the group proposed to work towards a taxonomy for software-based
contributorship.
The taxonomy should be derived from, or extend, existing taxonomies for
research impact and contributorship such as defined by CASRAI (in particular
based on the Wellcome-Harvard contributorship taxonomy\katznote{Dan to add cite or url}), VIVO, or ISNI.
An interesting measure of impact raised by the group was the {\em betweenness
centrality}, an indicator of a person's centrality (and hence, importance)
in a scientific collaboration.
It is expected that developers of research software often play such a central
role.

The group defined the following actions to be performed:
\begin{enumerate}
\item Investigate existing taxonomies for roles and contributorships.
\item Investigate prototype badging initiatives.
\item Investigate journals focusing on publishing peer-reviewed research
software.
\item Investigate guidelines and checklists of best practices.
\item Communicate the results of the above investigations to the WSSSPE
community, and to decision-making bodies (funders, publishers, universities,
and tenure committee representatives).
\item Ensure engagement of the broader research community in this discussion.
\end{enumerate}
